---
title: "SingleMoleculeGenomicsIO"
output:
    rmarkdown::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{SingleMoleculeGenomicsIO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`SingleMoleculeGenomicsIO` provides tools for reading and processing 
single-molecule genomics data in R. These include functions for reading base 
modifications from `modBam` files, e.g. generated by 
[Dorado](https://github.com/nanoporetech/dorado),
or text files generated by [modkit](https://nanoporetech.github.io/modkit),
efficient representation of such data as R objects, and functions to manipulate
and summarize such objects.

Current contributors include:

- [Panagiotis Papasaikas](https://github.com/ppapasaikas)
- [Sebastien Smallwood](https://github.com/smalseba)
- [Charlotte Soneson](https://github.com/csoneson)
- [Michael Stadler](https://github.com/mbstadler)

# Installation

`SingleMoleculeGenomicsIO` can be installed from GitHub via the 
`BiocManager` package:

```
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("fmicompbio/SingleMoleculeGenomicsIO")
```

# Load data

## Load read-level data

We start by loading single-molecule genomics data at the level of individual
reads (a small example dataset with 6mA modifications representing
accessibility). Read-level data is provided as a `modBam`
file (standard bam file with modification data stored in `MM` and `ML` tags,
see [SAMtags.pdf](https://samtools.github.io/hts-specs/SAMtags.pdf)) and is
read using the `readModBam()` function from `SingleMoleculeGenomicsIO`.

Alternatively, read-level modification data can also be extracted from
`modBam` files using [modkit](https://nanoporetech.github.io/modkit) (see 
`readModkitExtract()` to read the output generated by `modkit extract`).

The small example `modBam` files contained in `SingleMoleculeGenomicsIO` were
generated using the [Dorado](https://github.com/nanoporetech/dorado) aligner:

```{r data-files, message=FALSE}
# load packages
library(SingleMoleculeGenomicsIO)
library(SummarizedExperiment)
library(Biostrings)
library(SparseArray)

# read-level 6mA data generated by 'dorado'
modbamfiles <- system.file("extdata",
                           c("6mA_1_10reads.bam", "6mA_2_10reads.bam"),
                            package = "SingleMoleculeGenomicsIO")
names(modbamfiles) <- c("sample1", "sample2")
```

Modification data is read from these files using `readModBam()`:

```{r read-data}
se <- readModBam(bamfiles = modbamfiles,
                 regions = "chr1:6940000-7000000",
                 modbase = "a",
                 verbose = TRUE, 
                 BPPARAM = BiocParallel::SerialParam())
se
```

This will create a `RangedSummarizedExperiment` object with positions in rows:

```{r row-data}
# rows are positions...
rowRanges(se)
```

The columns of `se` correspond to samples:

```{r column-data}
# ... and columns are samples
colData(se)
```

The sample names in the `sample` column are obtained from the input files
(here `modbamfiles`), or if the files are not named will be automatically
assigned (using `"s1"`, `"s2"` and so on, assuming that each file corresponds
to a separate sample).

Read-level information is contained in a list under `readInfo`, with one
element per sample:

```{r coldata-read-info}
se$readInfo
```

Each sample typically contains several reads (see `se$n_reads`),
which correspond to the rows of `se$readInfo`, for example for `"sample1"`:

```{r coldata-read-info-sample1}
se$readInfo$sample1
```

## Explore assay data

The single assay `mod_prob` is a `DataFrame` with modification probabilities.

```{r mod-prob-assay}
assayNames(se)

m <- assay(se, "mod_prob")
m
```

In order to store read-level data for variable numbers of reads per sample,
each sample column is not just a simple vector, but a position-by-read
`NaMatrix` (a sparse matrix object defined in the
[`SparseArray`](https://bioconductor.org/packages/SparseArray) package).

```{r mod-prob-columns1}
m$sample1
```

If a 'flat' matrix is needed, in which columns correspond to
reads instead of samples, it can be created using `as.matrix()`.
Notice the number of columns in each of the objects:

```{r mod-prob-columns2}
ncol(se)           # 2 samples
ncol(m$sample1)    # 3 reads in "sample1"
ncol(m$sample2)    # 2 reads in "sample2"
ncol(as.matrix(m)) # 5 reads in total
```

One advantage of the grouping of reads per sample is that you can easily
perform per-sample operations using `lapply` (returns a `list`) or `endoapply`
(returns a `DataFrame`):

```{r mod-prob-lapply}
lapply(m, ncol)
endoapply(m, ncol)
```

The `NaMatrix` objects do not store `NA` values and thus use much less
memory compared to a normal (dense) matrix. The `NA` values here occur
at positions (rows) that are not covered by the read of that column, and there
are typically a large fraction of NA values:

```{r nnacount}
prod(dim(m$sample1)) # total number of values
nnacount(m$sample1)  # number of non-NA values
```

**Important: ** These `NA` values have to be excluded for example when
calculating the average modification probability for a position
using `mean(..., na.rm = TRUE)`, otherwise the result will be `NA` for
all incompletely covered positions:

```{r mod-prob-zeros}
# modification probabilities at position "chr1:6928850:-"
m["chr1:6928850:-", ]

# WRONG: take the mean of all values (including NAs)
lapply(m["chr1:6928850:-", ], mean)

# CORRECT: exclude the NA values (na.rm = TRUE)
lapply(m["chr1:6928850:-", ], mean, na.rm = TRUE)
```

Note that for sample2, in which no read overlaps the selected position,
`mean(..., na.rm = TRUE)` returns `NaN`.

In most cases it is preferable to use the convenience functions in
`SingleMoleculeGenomicsIO`, such as `flattenReadLevelAssay()` (see next
section), to calculate summaries over reads instead of the `lapply(...)` used
above for illustration.

## Summarize read-level data

Summarized data can be obtained from the read-level data by calculating
a per-position summary over the reads in each sample:

```{r summarise-read-level}
se_summary <- flattenReadLevelAssay(
    se, statistics = c("Nmod", "Nvalid", "FracMod", "Pmod", "AvgConf"))
```

As discussed above, this will correctly exclude non-observed values
and return zero values (for count assays `"Nmod"`, "`Nvalid`") or `NaN` for
assays with fractions (`"FracMod"`, `"Pmod"` and `"AvgConf"`), for positions
without any observed data (see for example `"Pmod"` in `"sample2"`):

```{r summarized-pmod}
assay(se_summary, "Pmod")["chr1:6928850:-", ]
```

The summary statistics to calculate are selected using the `statistics` argument.
By default, `flattenReadLevelAssay()` will count the number of modified (`Nmod`)
and total (`Nvalid`) reads at each position and sample, and calculate the
fraction of modified bases from the two (`FracMod`).

```{r default-statistics}
assay(se_summary, "Nmod")["chr1:6928850:-", ]
assay(se_summary, "Nvalid")["chr1:6928850:-", ]
assay(se_summary, "FracMod")["chr1:6928850:-", ]
```

In the above example, we additionally calculate the average modification
probability (`Pmod`) and the average confidence of the modification calls
per position (`AvgConf`). As we have not changed the default `keepReads = TRUE`,
we keep the read-level assay from the input object (`mod_prob`) in the returned
object. Because of the grouping of reads by sample in `mod_prob`, the dimensions
of read-level or summary-level objects remain identical:

```{r collapsed-assays}
# read-level data is retained in "mod_prob" assay
assayNames(se_summary)

# ... which groups the reads by sample
assay(se_summary, "mod_prob")

# the dimensions of  read-level `se` and summarized `se_summary` are identical
dim(se)
dim(se_summary)
```


## Load summary-level data

As illustrated above, summary-level data (e.g. average modification fraction at
each position) can be calculated from read-level data. Alternatively, it can
also be directly read from `modBam` files, which is faster and reduces the
size of the object in memory. In addition, there are other file formats that
only store summary-level data (e.g. `bed` files) that can also be imported
using `SingleMoleculeGenomicsIO`.

### Read summary-level data from `modBam` files

To read summary-level data with `readModBam()`, we set the `level` argument
to `"summary"`:

```{r read-data-summary}
se_summary2 <- readModBam(bamfiles = modbamfiles,
                          regions = "chr1:6940000-7000000",
                          modbase = "a",
                          level = "summary",
                          verbose = TRUE, 
                          BPPARAM = BiocParallel::SerialParam())
se_summary2
```

Notice that `se_summary2` contains only summary-level assays (`Nmod`, `Nvalid` and
`FracMod`), and no read-level modifications anymore (assay `mod_prob`).
Otherwise, it resembles the read-level object `se`, for example it covers the
same genomic positions:

```{r row-data-summary}
# rows are positions...
rowRanges(se)
rowRanges(se_summary2)
```

The summary-level assays generated from `se` (see `se_summary` above) are also
identical to the ones directly read from the `modBam` file into `se_summary2`:

```{r identical-summaries}
identical(assay(se_summary, "Nvalid"),
          assay(se_summary2, "Nvalid"))
```

### Read summary-level data from `bedMethyl` files

As mentioned, summary-level data can also be imported from `bedMethyl` files
that contain total and methylated counts for individual positions::

```{r data-files-summary}
# collapsed 5mC data for a small genomic window in bedMethyl format
bedmethylfiles <- system.file("extdata",
                              c("modkit_pileup_5mC_1.bed.gz",
                                "modkit_pileup_5mC_2.bed.gz"),
                              package = "SingleMoleculeGenomicsIO")

# file with sequence of the reference genome in fasta format
reffile <- system.file("extdata", "reference.fa.gz",
                       package = "SingleMoleculeGenomicsIO")
```

As these `bedMethyl` files represent 5mC modifications, we set `modbase = "m"`
in the `readBedMethyl()` call (the same could be done in `readModBam` to load
`modBam` files representing 5mC modifications). For a full list of
possible modifications see the section about base modifications in the
[SAM tag specification](https://samtools.github.io/hts-specs/SAMtags.pdf)). 
We also provide a sequence reference (genome sequence) from which to extract 
the sequence context for each position.

```{r read-data-bedmethyl}
names(bedmethylfiles) <- c("sample1", "sample2")
se_bed <- readBedMethyl(bedmethylfiles,
                        modbase = "m",
                        sequenceContextWidth = 3,
                        sequenceReference = reffile, 
                        BPPARAM = BiocParallel::SerialParam())
se_bed
```

As before, the summarized data is read into a `RangedSummarizedExperiment` 
object, with rows corresponding to genomic positions:

```{r rowranges-bedmethyl}
rowRanges(se_bed)
```

... and the columns correspond to the different samples (here corresponding
to the names of the two input files in `bedmethylfiles`):

```{r colnames-bedmethyl}
colnames(se_bed)
```

The data is stored in two matrices (assays) called `Nmod` (the number of
modified bases per sample and position) and `Nvalid` (the number of valid
total read covering that position in each sample):

```{r assays-bedmethyl}
assayNames(se_bed)

head(assay(se_bed, "Nmod"))
head(assay(se_bed, "Nvalid"))
```

From these, we can easily calculate the fraction of modifications:

```{r fraction-modified-bedmethyl}
fraction_modified <- assay(se_bed, "Nmod") / assay(se_bed, "Nvalid")
head(fraction_modified)
```

As before, non-finite fractions result from positions that had zero coverage 
in a given sample.

## Load data for a random subset of the reads

By default, `readModBam()` will load all reads overlapping the regions 
indicated by the `regions` argument. However, it is also possible to load a 
random subset of reads (this can be useful e.g. to calculate distributions 
of quality scores, using reads aligning to different parts of the genome but 
without the need to load the entire `modBam` file into R). 

```{r sample-reads}
se_sample <- readModBam(bamfiles = modbamfiles, 
                        modbase = "a", 
                        nAlnsToSample = 5, 
                        seqnamesToSampleFrom = "chr1", 
                        verbose = TRUE, 
                        BPPARAM = BiocParallel::SerialParam(RNGseed = 1327828L))
se_sample$n_reads
```

Note that the exact number of sampled reads may not be identical to the 
requested number. One reason for this is that the requested number is combined
with the total number of alignments to calculate the fraction of reads to be 
sampled (in this small case, 0.5), and each alignment will be retained with 
this probability. 

# Expand the `SummarizedExperiment` object to single base resolution

The objects returned by the reading functions above contain data only for 
the positions with an observable modification state for the given modbase. For
cases where a single nucleotide resolution is required (i.e., where the rows 
in the object must cover each position within a region), the 
`expandSEToBaseSpace` will return a suitable object. Note that any strand 
information in the original positions will be ignored.

```{r expand-se}
se_expand <- expandSEToBaseSpace(se)
dim(se_expand)
```


# Add quality scores and filter `SummarizedExperiment` object

`SingleMoleculeGenomicsIO` provides functionality to calculate a variety of 
statistics for individual reads. The results will be added as a new column in 
the `colData` of the `SummarizedExperiment`. 

```{r add-read-stats}
# start from read-level data
# ... some read-level statistics were added already when reading the data
se$readInfo$sample1

# ... calculate additional read statistics
se <- addReadStats(se, name = "QC")
se$QC$sample1
```

The distribution of the quality statistics can also be plotted, which is often 
helpful in order to decide on cutoffs for downstream analysis. Note that in the 
current example data, each sample contains only a few reads and hence the plot
below is very sparse. 

```{r plot-read-stats}
plotReadStats(se, readInfoCol = "readInfo", qcCol = "QC")
```

The `filterReads()` function can be used to filter the set of reads based on 
various quality indicators. This step requires that the read statistics that 
will be used for the filtering have been calculated as indicated above.

```{r filter-reads}
se_filtreads <- filterReads(se, readInfoCol = "readInfo",
                            qcCol = "QC", minQscore = 10, 
                            minAlignedFraction = 0.8)
# number of retained reads for each sample
lapply(assay(se_filtreads, "mod_prob"), ncol)
# tables with the reason behind each filtered read
metadata(se_filtreads)$filteredOutReads
```

It is also possible to filter out positions (rows), based on a variety of 
criteria, using the `filterPositions()` function. Typical filter criteria 
include the sequence context of the observation and the read coverage.

In order to filter by sequence context, we first need to add it to our 
object (if this was not already done when the data was imported). This can be 
done using the `addSeqContext()` function. We can extract sequence contexts of 
arbitrary width, based on a provided genome reference.

```{r add-seqcontext}
gnm <- readDNAStringSet(system.file("extdata",
                                    "reference.fa.gz",
                                    package = "SingleMoleculeGenomicsIO"))
se <- addSeqContext(se, sequenceContextWidth = 1, sequenceReference = gnm)
```

Next, we can filter the object to only contain positions where the annotated 
nucleotide at the corresponding genomic position is an `A`. Any IUPAC code can 
be used - for example, `NCG` would match trinucleotide contexts with a 
central C followed by a G (and preceded by any base).

```{r filter-seqcontext}
# check sequence context distribution across observations
table(as.character(rowData(se)$sequenceContext))

se_posfilt <- filterPositions(se, filters = "sequenceContext", 
                              sequenceContext = "A")
dim(se_posfilt)
# check that the sequence context of all retained positions is A
table(as.character(rowData(se_posfilt)$sequenceContext))
```

# Subset and regroup reads

After importing read-level data with `SingleMoleculeGenomicsIO`, the set of 
reads can be subset using the `subsetReads()` function. The subsetting can be 
done either by specifying the read names or indices to retain, or by asking 
for a random subset of reads. 

```{r subset-reads}
# subset to only the first two reads per sample
(reads_to_keep <- lapply(assay(se, "mod_prob"), function(s) 1:2))
se_subset <- subsetReads(se, reads = reads_to_keep)
lapply(assay(se_subset, "mod_prob"), ncol)

# subset to specified reads
(reads_to_keep <- unlist(lapply(assay(se, "mod_prob"), function(s) colnames(s)[c(1, 2)])))
se_subset <- subsetReads(se, reads = reads_to_keep)
lapply(assay(se_subset, "mod_prob"), ncol)

# request a random set of two reads per sample
se_subset <- subsetReads(se, randomSubset = 2)
lapply(assay(se_subset, "mod_prob"), ncol)
```

See `?subsetReads` for more subsetting options. 

In some situations, a _regrouping_ of the reads is more suitable than a 
subsetting. By default, the reads will be grouped by the sample they originate 
from (i.e., each column in the `SummarizedExperiment` corresponds to a sample). 
The `regroupReads()` and `regroupReadsByColData()` functions allow us to 
reorganize the reads in such a way that each column of the object instead 
corresponds to another read property (e.g., the strand from which the 
read originates). The new groups can be defined manually as a named list 
of read identifiers (for `regroupReads()`), or be generated automatically from 
a categorical column in the read-level column annotations (for 
`regroupReadsByColData()`). Here we illustrate how to regroup the reads by 
the strand. 

```{r regroup-reads}
# read strand is stored in the column annotations (ref_strand)
se$readInfo$sample1

# regroup reads by strand, ignoring the sample information
se_regroup <- regroupReadsByColData(se, colNames = "ref_strand", 
                                    withinSample = FALSE)
colnames(se_regroup)

# regroup reads by strand, within each sample
se_regroup <- regroupReadsByColData(se, colNames = "ref_strand", 
                                    withinSample = TRUE)
colnames(se_regroup)
```

# Add read segment annotations

Read segment annotation allow us to store information about specific subregions
of reads. These could, for example, represent inferred locations of nucleosomes
or transcription factor binding sites. The segments need to be provided as a 
list of `IRangesList` objects (one per sample), with each element of an 
`IRangesList` providing the annotations for one read. 

```{r annotate-segments}
# define segments
segs <- list(sample1 = IRangesList(
    "sample1-d52a5f6a-a60a-4f85-913e-eada84bfbfb9" = IRanges(
        start = 6925834, width = 140),
    "sample1-233e48a7-f379-4dcf-9270-958231125563" = IRanges(
        start = c(6926000, 6926200), width = 140)), 
    sample2 = IRangesList())

# add segment annotation to se
se <- annotateReadSegments(se, irlList = segs, name = "nucl")
se$nucl
```

# Count state pairs

The import functions exemplified above all return a `SummarizedExperiment` 
object with read- and/or summary-level data. For certain applications, 
other data representations are more suitable. The `countStatePairs()` 
function allows us to parse a `modBam` file and tabulate the number of pairs 
of bases at a given distance and with a given combination of modification 
states. 

```{r count-state-pairs}
sp <- countStatePairs(bamfile = modbamfiles[1], 
                      modbase = "a", 
                      windowSize = 200, 
                      BPPARAM = BiocParallel::SerialParam(), 
                      verbose = FALSE)
sp
```


# Session info

```{r session-info}
sessioninfo::session_info()
```

