---
title: "SingleMoleculeGenomicsIO"
output:
    rmarkdown::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{SingleMoleculeGenomicsIO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`SingleMoleculeGenomicsIO` provides tools for reading single-molecule genomics
data in R. These include functions for reading base modifications from `modBam`
files, e.g. generated by [Dorado](https://github.com/nanoporetech/dorado),
or text files generated by [modkit](https://nanoporetech.github.io/modkit),
efficient representation of such data as R objects, and functions to manipulate
and summarize such objects.

Current contributors include:

- [Panagiotis Papasaikas](https://github.com/ppapasaikas)
- [Sebastien Smallwood](https://github.com/smalseba)
- [Charlotte Soneson](https://github.com/csoneson)
- [Michael Stadler](https://github.com/mbstadler)

## Installation

`SingleMoleculeGenomicsIO` can be installed from GitHub via the 
`BiocManager` package:

```
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("fmicompbio/SingleMoleculeGenomicsIO")
```

## Load read-level data

We start by loading single-molecule genomics data at the level of individual
reads (a small example dataset with 6mA modifications representing
accessibility). Read-level data is provided as a `modBam`
file (standard bam file with modification data stored in `MM` and `ML` tags,
see [SAMtags.pdf](https://samtools.github.io/hts-specs/SAMtags.pdf)) and is
read using the `readModBam()` function from `SingleMoleculeGenomicsIO`.

Alternatively, read-level modification data can also be extracted from
`modBam` files using [modkit](https://nanoporetech.github.io/modkit) (see 
`readModkitExtract()` to read the output generated by `modkit extract`).

The small example `modBam` files contained in `SingleMoleculeGenomicsIO` were
generated using the [Dorado](https://github.com/nanoporetech/dorado) aligner:

```{r data-files, message=FALSE}
# load packages
library(SingleMoleculeGenomicsIO)
library(SummarizedExperiment)
library(SparseArray)

# read-level 6mA data generated by 'dorado'
modbamfiles <- system.file("extdata",
                           c("6mA_1_10reads.bam", "6mA_2_10reads.bam"),
                            package = "SingleMoleculeGenomicsIO")
names(modbamfiles) <- c("sample1", "sample2")
```

Modification data is read from these files using `readModBam()`:

```{r read-data}
se <- readModBam(bamfiles = modbamfiles,
                 regions = "chr1:6940000-7000000",
                 modbase = "a",
                 verbose = TRUE, 
                 BPPARAM = BiocParallel::SerialParam())
se
```

This will create a `RangedSummarizedExperiment` object with positions in rows:

```{r row-data}
# rows are positions...
rowRanges(se)
```

The columns of `se` correspond to samples:

```{r column-data}
# ... and columns are samples
colData(se)
```

The sample names in the `sample` column are obtained from the input files
(here `modbamfiles`), or if the files are not named will be automatically
assigned (using `"s1"`, `"s2"` and so on, assuming that each file corresponds
to a separate sample).

Read-level information is contained in a list under `readInfo`, with one
element per sample:

```{r coldata-read-info}
se$readInfo
```

Each sample typically contains several reads (see `se$n_reads`),
which correspond to the rows of `se$readInfo`, for example for `"sample1"`:

```{r coldata-read-info-sample1}
se$readInfo$sample1
```

## Explore assay data

The single assay `mod_prob` is a `DataFrame` with modification probabilities.

```{r mod-prob-assay}
assayNames(se)

m <- assay(se, "mod_prob")
m
```

In order to store read-level data for variable numbers of reads per sample,
each sample column is not just a simple vector, but a position-by-read
`NaMatrix` (a sparse matrix object defined in the
[`SparseArray`](https://bioconductor.org/packages/SparseArray) package).

```{r mod-prob-columns1}
m$sample1
```

If a 'flat' matrix is needed, in which columns correspond to
reads instead of samples, it can be created using `as.matrix()`.
Notice the number of columns in each of the objects:

```{r mod-prob-columns2}
ncol(se)           # 2 samples
ncol(m$sample1)    # 3 reads in "sample1"
ncol(m$sample2)    # 2 reads in "sample2"
ncol(as.matrix(m)) # 5 reads in total
```

One advantage of the grouping of reads per sample is that you can easily
perform per-sample operations using `lapply` (returns a `list`) or `endoapply`
(returns a `DataFrame`):

```{r mod-prob-lapply}
lapply(m, ncol)
endoapply(m, ncol)
```

The `NaMatrix` objects do not store `NA` values and thus use much less
memory compared to a normal (dense) matrix. The `NA` values here occur
at positions (rows) that are not covered by the read of that column, and there
are typically a large fraction of NA values:

```{r nnacount}
prod(dim(m$sample1)) # total number of values
nnacount(m$sample1)  # number of non-NA values
```

**Important: ** These `NA` values have to be excluded for example when
calculating the average modification probability for a position
using `mean(..., na.rm = TRUE)`, otherwise the result will be `NA` for
all incompletely covered positions:

```{r mod-prob-zeros}
# modification probabilities at position "chr1:6928850:-"
m["chr1:6928850:-", ]

# WRONG: take the mean of all values (including NAs)
lapply(m["chr1:6928850:-", ], mean)

# CORRECT: exclude the NA values (na.rm = TRUE)
lapply(m["chr1:6928850:-", ], mean, na.rm = TRUE)
```

Note that for sample2, in which no read overlaps the selected position,
`mean(..., na.rm = TRUE)` returns `NaN`.

In most cases it is preferable to use the convenience functions in
`SingleMoleculeGenomicsIO`, such as `flattenReadLevelAssay()` (see next
section), to calculate summaries over reads instead of the `lapply(...)` used
above for illustration.

## Summarize read-level data

Summarized data can be obtained from the read-level data by calculating
a per-position summary over the reads in each sample:

```{r summarise-read-level}
se_summary <- flattenReadLevelAssay(
    se, statistics = c("Nmod", "Nvalid", "FracMod", "Pmod", "AvgConf"))
```

As discussed above, this will correctly exclude non-observed values
and return zero values (for count assays `"Nmod"`, "`Nvalid`") or `NaN` for
assays with fractions (`"FracMod"`, `"Pmod"` and `"AvgConf"`), for positions
without any observed data (see for example `"Pmod"` in `"sample2"`):

```{r summarized-pmod}
assay(se_summary, "Pmod")["chr1:6928850:-", ]
```

The summary statistics to calculate are selected using the `statistics` argument.
By default, `flattenReadLevelAssay()` will count the number of modified (`Nmod`)
and total (`Nvalid`) reads at each position and sample, and calculate the
fraction of modified bases from the two (`FracMod`).

```{r default-statistics}
assay(se_summary, "Nmod")["chr1:6928850:-", ]
assay(se_summary, "Nvalid")["chr1:6928850:-", ]
assay(se_summary, "FracMod")["chr1:6928850:-", ]
```

In the above example, we additionally calculate the average modification
probability (`Pmod`) and the average confidence of the modification calls
per position (`AvgConf`). As we have not changed the default `keepReads = TRUE`,
we keep the read-level assay from the input object (`mod_prob`) in the returned
object. Because of the grouping of reads by sample in `mod_prob`, the dimensions
of read-level or summary-level objects remain identical:

```{r collapsed-assays}
# read-level data is retained in "mod_prob" assay
assayNames(se_summary)

# ... which groups the reads by sample
assay(se_summary, "mod_prob")

# the dimensions of  read-level `se` and summarized `se_summary` are identical
dim(se)
dim(se_summary)
```


## Load summary-level data

As illustrated above, summary-level data (e.g. average modification fraction at
each position) can be calculated from read-level data. Alternatively, it can
also be directly read from `modBam` files, which is faster and reduces the
size of the object in memory. In addition, there are other file formats that
only store summary-level data (e.g. `bed` files) that can also be imported
using `SingleMoleculeGenomicsIO`.

### Read summary-level data from `modBam` files

To read summary-level data with `readModBam()`, we set the `level` argument
to `"summary"`:

```{r read-data-summary}
se_summary2 <- readModBam(bamfiles = modbamfiles,
                          regions = "chr1:6940000-7000000",
                          modbase = "a",
                          level = "summary",
                          verbose = TRUE, 
                          BPPARAM = BiocParallel::SerialParam())
se_summary2
```

Notice that `se_summary2` contains only summary-level assays (`Nmod`, `Nvalid` and
`FracMod`), and no read-level modifications anymore (assay `mod_prob`).
Otherwise, it resembles the read-level object `se`, for example it covers the
same genomic positions:

```{r row-data-summary}
# rows are positions...
rowRanges(se)
rowRanges(se_summary2)
```

The summary-level assays generated from `se` (see `se_summary` above) are also
identical to the ones directly read from the `modBam` file into `se_summary2`:

```{r identical-summaries}
identical(assay(se_summary, "Nvalid"),
          assay(se_summary2, "Nvalid"))
```

### Read summary-level data from `bed` files

As mentioned, summary-level data can also be imported from `bedMethyl` files
that contain total and methylated counts for individual positions::

```{r data-files-summary}
# collapsed 5mC data for a small genomic window in bedMethyl format
bedmethylfiles <- system.file("extdata",
                              c("modkit_pileup_5mC_1.bed.gz",
                                "modkit_pileup_5mC_2.bed.gz"),
                              package = "SingleMoleculeGenomicsIO")

# file with sequence of the reference genome in fasta format
reffile <- system.file("extdata", "reference.fa.gz",
                       package = "SingleMoleculeGenomicsIO")
```

### WAS HERE

This allows us to read the data using `readBedMethyl()` (`modbase = "m"`
indicates that we want to read 5mC modifications, for a full list of
possible modifications see the section about base modifications in the
[SAM tag specification](https://samtools.github.io/hts-specs/SAMtags.pdf)):

```{r read-data}
names(bedmethylfiles) <- c("sample1", "sample2")
se <- readBedMethyl(bedmethylfiles,
                    modbase = "m",
                    sequenceContextWidth = 3,
                    sequenceReference = reffile, 
                    BPPARAM = BiocParallel::SerialParam())
se
```

These `bedMethyl` files contain summarized data and are read into a
`RangedSummarizedExperiment` object. The rows here correspond to genomic
positions:

```{r rowranges}
rowRanges(se)
```

... and the columns correspond to the different samples (here corresponding
to the names of the two input files in `bedmethylfiles`):

```{r colnames}
colnames(se)
```

The data is stored in two matrices (assays) called `Nmod` (the number of
modified bases per sample and position) and `Nvalid` (the number of valid
total read covering that position in each sample):

```{r assays}
assayNames(se)

head(assay(se, "Nmod"))
head(assay(se, "Nvalid"))
```

From these, you can easily calculate the fraction of modifications:

```{r fraction-modified}
fraction_modified <- assay(se, "Nmod") / assay(se, "Nvalid")
head(fraction_modified)
```

Please note that non-finite fractions result from positions that had
zero coverage in a given sample.

### Plot data

The `plotRegion()` function provides a convenient and flexible interface for 
visualizing data for a specific region. Here we visualize the accessibility 
using GpC methylation (sequence context GCH), and adding a smooth line:

```{r plot-region-GCH, warning=FALSE, fig.width=6, fig.height=4, fig.alt="A point plot showing the fraction of modified bases per position in GCH contexts, colored by sample."}
plotRegion(se,
           region = "chr1:6939000-6946000",
           tracks = list(list(trackData = "FracMod",
                              trackType = "PointSmooth")),
           sequenceContext = "GCH")
```

We can also restrict the plotted sites to focus on endogenous DNA methylation 
(sequence context HCG):

```{r plot-region-HCG, warning=FALSE, fig.width=6, fig.height=4, fig.alt="A point plot showing the fraction of modified bases per position in HCG contexts, colored by sample."}
plotRegion(se,
           region = "chr1:6939000-6946000",
           tracks = list(list(trackData = "FracMod",
                              trackType = "PointSmooth")),
           sequenceContext = "HCG")
```

Or just plot the coverage (`Nvalid`) as a smooth line:

```{r plot-region-Nvalid, warning=FALSE, fig.width=6, fig.height=4, fig.alt="A smoothed line plot showing the read coverage (Nvalid) per position, colored by sample."}
plotRegion(se,
           region = "chr1:6939000-6946000",
           tracks = list(list(trackData = "Nvalid",
                              trackType = "Smooth")))
```

## Session info

```{r session-info}
sessioninfo::session_info()
```

